---
title: "scot-cats"
author: "Eric C. Anderson"
date: "July 2, 2015"
output: html_document
---

```{r, setup, echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "~/Desktop/scot-cats/PlainPars")
opts_chunk$set(comment = NA)
```


This document shows how to use the R package `Rrunstruct` to use R to conduct multiple `structure`
runs based on the parameters chosen for any parameter set specified in the `structure` front-end.
It goes hand in hand with the computer practical in the SISG MCMC course. Before you follow this
document, you will want to have completed the first part of the computer practical and have done
at least one run of `structure` using the front-end.

The goal here is to see how to use some R tools to parse `structure` output to be able to quickly
assess whether multiple runs of `structure's` MCMC algorithm give similar results.  Along the way
we will encounter a couple of challenges that arise, most notably the issue that the labels of
the clusters may not be concordant across multiple `structure` runs.  We might also take some time
to look at some of the tools in today's Hadley Wickham R ecosystem (`dplyr`, `ggplot2`, `tidyr`, `stringr`, etc.) that are helpful.

Before we can begin there is some setup you might have to do.  


## Setup
You will need to install the package `Rrunstruct`.  Because the package includes the `structure`
executable file, it is not available on CRAN, but can be installed from GitHub.  Doing so
requires the `devtools` package.
```{r, eval=FALSE}
devtools::install_github("eriqande/Rrunstruct")
```
If that doesn't work, it likely is because you need to get the `devtools` package and then try again:
```{r, eval=FALSE}
installed.packages("devtools")
devtools::install_github("eriqande/Rrunstruct")
```




## Running Structure
To run `structure` you just need to:

1. load the package
2. set your R working directory to the directory inside the `structure` project that holds the
parameter set you want to use.  On my computer that is `~/Desktop/scot-cats/PlainPars`.  On 
your computer, that path will have to be set accordingly.
3. Use the function`structure_runs()`, telling it where to put the output, what K values to use, how
many reps for each, and the seed to use for the first run (this gets incremented by one for each
successive run).
```{r, message=FALSE}
library(Rrunstruct)
```
```{r}
setwd("~/Desktop/scot-cats/PlainPars")  # you've got to change this for your own computer!
```

```{r, eval=FALSE}
structure_runs(outdir = "struct-runs-from-r-1", K = 1:4, reps = 3, seed = 5566)
```

Once those runs have completed, you can take a look at what the outputs look like.  All the output
files should be in the directory `struct-runs-from-r-1` that is inside the `PlainPars` directory.
Let's see what they look like:
```{r}
dir(path = "struct-runs-from-r-1")
```
The naming convention is pretty simple, `*results*` are the files that structure spits out at
the end of each run.  The `*stdout*` files store the stuff that the command line version of 
structure spits out.  Let's have a look at the first couple lines of the trace files:
```{r}
cat(readLines("struct-runs-from-r-1/struct_stdout_K_3_rep_1")[1:50], sep = "\n")
```

And look at a bit of one of the result files.  Note that you are going to want to extract
the "Inferred ancestry of individuals from this file"
```{r, warning=FALSE}
cat(readLines("struct-runs-from-r-1/struct_results_K_3_rep_1_f")[1:100], sep = "\n")
```


## Slurping up the Structure results

Although R's syntax for text processing is perhaps not as compact as `awk` or `sed`, and it might be slower
than `python` it has nice facilities for parsing text files.  The package `stringr` used in combination with
the function `readLines()` is particularly easy to use and helpful.  

We won't go into the details now though.  Instead we have a single function that:

1.  Reads in all the `stdout` and `results` files and wrangles them into long-format (tidy) data frames.
2. Identifies the cluster that each individual has the highest mean posterior Q for from each run.
3. For each rep determine the permutation of the cluster labels that makes the results the most congruent with
the first rep.  (This is a lightweight version of what the program CLUMPP does)
4. Creates columns in the data frame that have permuted versions of the cluster labels that can be used in plotting.

If you want the details of all that, you can check the package out on GitHub.  I used the packages `tidyr` and `dplyr`
extensively (but the code is still kind of ugly...).  

Here we get all the output into a list of two data frames:
```{r, message=FALSE, cache=TRUE}
L <- read_and_process_structure_output("struct-runs-from-r-1")
```

And let's take a look at the tops of the two data frames in that list.  

### Q values
First, the estimated
Q values from each individual
```{r}
L$q
```
The columns `K` and `Rep` just give the K-value assumed in the structure run, and which replicate
the run was.  `Index` is just the numerical index of the individual, `Label` is the label given
to the individual sample, `Miss` is the percentage of loci missing data in that individual, `Pop`
is the putative popuation or sampling location of the individual (as given in the data input),
`cluster` gives the cluster that the `probability` refers to.  `probability` is the posterior
mean of the Q value for the given cluster.  `mp` is the number of the label permutation that makes
the results of this replicate as congruent as possible with replicate #1.  Finally
`cluster_relabeled` is the label of the cluster, once the clusters have been relabeled to be
congruent with replicate #1.  We will explore that further.


### Traces
Here we look at the traces:
```{r}
L$traces
```
The columns `K`, `Rep`, and `mp` are as above.  `Sweep` tells us which sweep the row refers
to.  It runs from 1 to 6000 in this case---it includes sweeps from the burn-in period. `variable`
tells us what variable we are looking at.  Here, those are either "Alpha" (the $\alpha$ variable)
or one of the $F$ parameters (i.e. F1, F2, etc).  `variables_relabeled` has the name of the
variable after they have been relabeled to be (more or less) congruent with replicate 1. `value` is
the value of that variable.

## Plotting the traces
Once structure's outputs are in this type of long-format data frame, it is relatively easy
to plot and compare between runs.  

### Alpha
First we will plot the traces of $\alpha$ and the $F$ parameters and see how the compare
across runs.  

Here are the $\alpha$'s:
```{r, fig.height=10, fig.width=14}
ggplot(L$traces %>% filter(variable == "Alpha"), aes(x = Sweep, y = value)) +
  geom_line(colour = "blue") +
  facet_grid(K ~ Rep)
```
That is interesting.  Not surprisingly, when K is 1, $\alpha$ meanders all over because
it is not really used in the model when K is 1, so there is nothing constraining it.  

For $K>1$ the results look pretty similar between the reps.  But, let's focus on $K>1$:
```{r, fig.height=10, fig.width=14}
ggplot(L$traces %>% filter(variable == "Alpha", K != 1), aes(x = Sweep, y = value)) +
  geom_line(colour = "blue") +
  facet_grid(K ~ Rep)
```

### F's
When we look at the values of F, we see the labeling issue:
```{r, fig.height=10, fig.width=14}
ggplot(L$traces %>% filter(str_detect(variable, "^F")), aes(x = Sweep, y = value, colour = variable)) +
  geom_line() +
  facet_grid(K ~ Rep)
```

When $K>1$ the colors don't match up because the clusters don't have the same name.  We can
rectify that by using the relabeled version:
```{r, fig.height=10, fig.width=14}
ggplot(L$traces %>% filter(str_detect(variables_relabeled, "^F")), aes(x = Sweep, y = value, colour = variables_relabeled)) +
  geom_line() +
  facet_grid(K ~ Rep)
```
We see that the course of the F variables during MCMC looks similar between reps for K=1 and K=2
but quite different between reps for K>2.  We contend that the runs for K>2 might have to be run
for much longer (and even then they may not converge to the same place, reliably).

## Plotting the Q values
Often what we are really interested in are the Q values for different individuals.  It is thus
important to assess how much those differ between runs.  One way to do that is to plot the 
estimated Q values in each rep > 1 with those in the first rep.  To do that, we can add the 
rep1 results (at each K) to the data frame of results:
```{r}
Qs <- L$q %>%
  filter(Rep == 1) %>%  # get just the first rep
  group_by(K, Index, cluster_relabeled) %>% # group by these to preserve the columns
  transmute(rep1_cluster = cluster_relabeled, rep1_prob = probability) %>% # rename columns
  ungroup %>%
  inner_join(L$q) %>%  # slap those values back on the original data frame matching K and Rep
  filter(Rep != 1) %>% # toss out the rows comparing Rep 1 to Rep 1
  mutate(which_cluster = paste(rep1_cluster))  # name a new column with something better for the plot
```
Here is what that data frame looks like:
```{r}
Qs
```
And now it is easy to compare the Q values in each Rep after Rep 1 with the Q values in Rep 1:
```{r, fig.height=10, fig.width=14}
ggplot(Qs, aes(x = rep1_prob, y = probability, colour = which_cluster)) +
  geom_point() +
  facet_grid(K ~ Rep)
```

This is interesting as it shows that for K=2 the three runs all converged to pretty much the
same Q values.  The "wiggle" off the $y=x$ line there is due to Monte Carlo variance, but not
caused from the chains converging to very different parts of the space, as is clear what has
happened for some of the replicates for K=3 and K=4.

## Longer runs at K = 2
Let's see if we can tighted up that Monte Carlo variance be doing longer runs.  I have made a
new directory `LongerRun` with the `structure` GUI and set it up to to 5000 burn in and 45000
sweeps after burn in.  Let's do 4 reps at K=2 to see how that works out.
```{r, eval=FALSE}
# do the runs
setwd("~/Desktop/scot-cats/LongerRun")  # this must be changed on other computers
structure_runs(outdir = "r-runs-2", K = 2, reps = 4, seed = 9876)
```
Then slurp up the results
```{r, cache = TRUE}
setwd("~/Desktop/scot-cats/LongerRun")  # this must be changed on other computers
L2 <- read_and_process_structure_output("r-runs-2")
```

Then, compare rep 1 to the others:
```{r}
# format the data frame
Qs2 <- L2$q %>%
  filter(Rep == 1) %>%  # get just the first rep
  group_by(K, Index, cluster_relabeled) %>% # group by these to preserve the columns
  transmute(rep1_cluster = cluster_relabeled, rep1_prob = probability) %>% # rename columns
  ungroup %>%
  inner_join(L2$q) %>%  # slap those values back on the original data frame matching K and Rep
  filter(Rep != 1) %>% # toss out the rows comparing Rep 1 to Rep 1
  mutate(which_cluster = paste(rep1_cluster))  # name a new column with something better for the plot
```

```{r}
# plot those up:
```{r, fig.height=5, fig.width=14}
ggplot(Qs2, aes(x = rep1_prob, y = probability, colour = which_cluster)) +
  geom_point() +
  facet_wrap( ~ Rep, ncol = 3)
```

Those look pretty concordant apart from a few individuals.