---
title: "Listing of Available R-code Companions"
output: html_document
---

This is a listing, by session, of all the R-code companions available for 
the SISG MCMC course, as well as a few other materials.

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(stringr)
```

```{r, echo=FALSE}
# get the name of all .nb.html files in docs, and the corresponding Rmd name
nbfull <- dir("../docs", pattern = "*.nb.html", full.names = TRUE) %>%
  setNames(dir("../docs", pattern = "*.nb.html")) %>%
  enframe(name = "nbfile", value = "nb_full") %>%
  mutate(file = str_replace(nbfile, "nb.html$", "Rmd"))


# make a tibble of all Rmd files in the R-notebooks directory and filter it by nb.htmls in docs
rmds <- dir("../R-notebooks", pattern = "*.Rmd", full.names = TRUE) %>%
  setNames(dir("../R-notebooks/", pattern = "*.Rmd")) %>%
  enframe(name = "file", value = "path") %>%
  semi_join(nbfull, by = "file")


# now, get the titles and descriptions and pull out the session and pages
rtib <- rmds %>%
  mutate(title = lapply(rmds$path, function(x) {
    tmp <- packagedocs::read_rmd_yaml(x)$title
    if (is.null(tmp)) tmp <- "No Title"
    tmp})) %>%
  mutate(description = lapply(rmds$path, function(x) {
    tmp <- packagedocs::read_rmd_yaml(x)$description
    if (is.null(tmp)) tmp <- "No Description"
    tmp}))  %>%
  mutate(session = parse_number(str_match(file, "s([0-9]+)")[,2]),
         session = ifelse(is.na(session), 0, session)) %>%
  arrange(session, file) %>%
  mutate(nbfile = str_replace(file, ".Rmd", ".nb.html")) %>%
  mutate(lstring = paste0("- [", title, "](./", nbfile, ").  ", description, "\n"))

```


```{r, echo = FALSE, results = "asis"}
# Now, we just break those into sessions and print them as markdown into the document
sess <- split(rtib, rtib$session)
dump <- lapply(names(sess), function(s) {
  cat(paste0("\n\n## Session ", s, "\n\n"))
  cat(sess[[s]]$lstring, sep = "")
})
```

## Data sets

- You can get the `cats.dat` data set by clicking [this link](https://raw.githubusercontent.com/eriqande/sisg_mcmc_course/master/computing_practicals/cats.dat), and 
then copying the data and putting it into a text file on your computer named `cats.dat`.

## Additional Lecture-like things

- [Thermodynamic Integration](./thermodynamic-integration.nb.html)
- [Matthew Stephens' 5-minute Statistics Page](http://stephens999.github.io/fiveMinuteStats/)
- [Monte Carlo Methods and Importance Sampling](http://ib.berkeley.edu/labs/slatkin/eriq/classes/guest_lect/mc_lecture_notes.pdf). A short thing that Eric wrote when a grad student and substitute lecturing.  This formed the basis for some of the material on vanilla Monte Carlo and Importance Sampling.
